### 2024-02-20 presentation/parts/introduction.tex
Work on presentation

\begin{frame}
    \frametitle{Documenting medical encounters}
    \begin{itemize}
        \item <1,2> Documentation is a central part of healthcare.
        \item <1,2> E.g. patient records, insurance claims, billing, research, training, legal purposes.
        \vspace{1em}
        \item <2> {\color{dtured}Time}: Physicians spend 34-37\% of their time on documentation \cite{joukes_time_2018, tipping_where_2010, sinsky_allocation_2016}\footnote{Ambulatory care across four specialties in four states and tertiary care at an academic medical center.}.
        \item <2> {\color{dtured}Quality}: Discharge summaries rarely meet all timeline, transmission, and content criteria. \cite{horwitz_comprehensive_2013}\footnote{Outpatient visits, Yale-New Haven Hospital.}
    \end{itemize}

>>>>

\begin{frame}
    \frametitle{Documenting medical encounters}
    \begin{itemize}
        \item <1,2> Documentation is a central part of healthcare.
        \item <1,2> E.g. patient records, insurance claims, billing, research, training, legal purposes.
        \vspace{1em}
        \item <2> {\color{dtured}Time-consuming}: Physicians spend 34-37\% of their time on documentation \cite{joukes_time_2018, tipping_where_2010, sinsky_allocation_2016}\footnote{Ambulatory care across four specialties in four states and tertiary care at an academic medical center.}.
        \item <2> {\color{dtured}Varying quality}: Discharge summaries rarely meet all timeline, transmission, and content criteria. \cite{horwitz_comprehensive_2013}\footnote{Outpatient visits, Yale-New Haven Hospital.}
    \end{itemize}

### 2024-02-20 presentation/parts/presentation_hierarchical.tex
More presentation work
Updates to presentation
WIP
Work on presentation

\section{Out-of-distribution detection}

@@

\frame{
    \frametitle{Problem and Contributions}
    \begin{itemize}
        \item Deep generative models often fail at OOD detection task when using their likelihood estimate as the score function \cite{nalisnick_deep_2019} by, perhaps surprisingly, assigning \textbf{higher likelihoods} to the OOD data.
        \item Contributions:
        \begin{itemize}
            \item We present a fast and fully unsupervised method for OOD detection competitive with the state-of-the-art
            \item We provide evidence that out-of-distribution detection fails due to learned low-level features that generalize across datasets.
        \end{itemize}  
    \end{itemize}
}

@@

\section{Models}

@@

\section{The Problem}

@@

\frame{
    \frametitle{Importance sampling the ELBO}
    The well-known importance weighted autoencoder (IWAE) bound is tight with the true likelihood in the limit of infinite samples, $S\rightarrow\infty$ \cite{burda_importance_2016},
    \begin{equation}\label{eq:iw-bound}
        \mathcal{L}_{S} = \mathbb{E}_{q(\zb|\xb)}\left[ \log \frac{1}{N} \sum_{s=1}^{S} \frac{p(\xb, \zb^{(s)})}{q(\zb^{(s)}|\xb)} \right] \leq \log p_\theta(\xb) \ ,
    \end{equation}

@@

    Consequently, by importance sampling the ELBO, the associated KL-divergence associated vanishes and our likelihood ratio reduces to the KL-divergence associated with $\mathcal{L}^{>k}$.

@@

    We can now see that $LLR^{>k}_{S}(\xb)$ performs OOD detection based on the top-most latent variables.
}

@@

\frame{
    \frametitle{Results with $LLR^{>k}$}
    The score has good performance across many different datasets.

    \begin{columns}

        \begin{table}[t]
            \centering
            \resizebox{0.9\textwidth}{!}{%
            \begin{tabular}{llrrr}
                \toprule
                 OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on CIFAR10}} \\
                 \midrule
        SVHN          &  $LLR^{>2}$  &  0.811  &  0.837  &  0.394  \\
        CIFAR10       &  $LLR^{>1}$  &  0.469  &  0.479  &  0.835  \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on SVHN}} \\
                 \midrule
        CIFAR10            &  $LLR^{>1}$       &  $0.939$  &  $0.950$  &  $0.052$  \\
        SVHN               &  $LLR^{>1}$       &  $0.489$  &  $0.484$  &  $0.799$  \\
            \bottomrule
            \end{tabular}
            }
        \end{table}
    \end{column}

    \begin{column}{0.5\textwidth}
        \begin{table}[t]
            \centering
            \resizebox{0.9\textwidth}{!}{%
            \begin{tabular}{llrrr}
                \toprule
                 OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on FashionMNIST}} \\
                 \midrule
        MNIST                    & $LLR^{>1}$             &  0.986  &  0.987  &  0.011 \\
        notMNIST                 &  $LLR^{>1}$            &  0.998  &  0.998  &  0.000 \\
        KMNIST                   &  $LLR^{>1}$            &  0.974  &  0.977  &  0.017 \\
        Omniglot28x28            &  $LLR^{>2}$            &  1.000  &  1.000  &  0.000 \\
        Omniglot28x28Inverted    &  $LLR^{>1}$            &  0.954  &  0.954  &  0.050 \\
        SmallNORB28x28           &  $LLR^{>2}$            &  0.999  &  0.999  &  0.002 \\
        SmallNORB28x28Inverted   &  $LLR^{>2}$            &  0.941  &  0.946  &  0.069 \\
        FashionMNIST             &  $LLR^{>1}$            &  0.488  &  0.496  &  0.811 \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on MNIST}} \\
                 \midrule
        FashionMNIST                   &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
        notMNIST                       &  $LLR^{>1}$  &  $1.000$  &  $0.999$  &  $0.000$ \\
        KMNIST                         &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
        Omniglot28x28                  &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
        Omniglot28x28Inverted          &  $LLR^{>1}$  &  $0.944$  &  $0.953$  &  $0.057$ \\
        SmallNORB28x28                 &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
        SmallNORB28x28Inverted         &  $LLR^{>1}$  &  $0.985$  &  $0.987$  &  $0.000$ \\
        MNIST                          &  $LLR^{>2}$  &  $0.515$  &  $0.507$  &  $0.792$ \\
                 \bottomrule
            \end{tabular}
            }
            % \caption{Additional results for the HVAE model trained on FashionMNIST. All results computed with 1000 importance samples.}
            % \label{tab:additional-results-fashionmnist}
        \end{table}
    \end{column}

>>>>

\section{Out-of-distribution detection}
% \subsection{Out-of-distribution detection}

@@

\frame{
    \frametitle{Problem and Contributions}
    \begin{itemize}
        \item Deep generative models often fail at OOD detection task when using their likelihood estimate as the score function \cite{nalisnick_deep_2019} by, perhaps surprisingly, assigning \textbf{higher likelihoods} to the OOD data.
        \item Contributions:
        \begin{itemize}
            \item We provide evidence that out-of-distribution detection fails due to learned low-level features that generalize across datasets.
            \item We present a fast and fully unsupervised method for OOD detection competitive with the state-of-the-art
        \end{itemize}  
    \end{itemize}
}

@@

\section{Latent variable models}

@@

\section{Identifying the issue}

@@

\frame{
    \frametitle{Importance sampling the ELBO}
    The importance weighted autoencoder (IWAE) bound is tight with the true likelihood in the limit of infinite samples, $S\rightarrow\infty$ \cite{burda_importance_2016},
    \begin{equation}\label{eq:iw-bound}
        \mathcal{L}_{S} = \mathbb{E}_{q(\zb|\xb)}\left[ \log \frac{1}{N} \sum_{s=1}^{S} \frac{p(\xb, \zb^{(s)})}{q(\zb^{(s)}|\xb)} \right] \leq \log p_\theta(\xb) \ ,
    \end{equation}

@@

    Consequently, by importance sampling the ELBO, the associated KL-divergence vanishes and our likelihood ratio reduces to the KL-divergence of $\mathcal{L}^{>k}$.

@@

    $LLR^{>k}_{S}(\xb)$ performs KL-divergence-based OOD detection using top-most latent variables.
}

@@

\begin{frame}
    \frametitle{Results on FashionMNIST/MNIST}
    \begin{table}
        \centering
        \resizebox*{!}{0.87\textheight}{%
        \begin{tabular}{lrrr}
            \toprule
             Method & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
             \midrule
             \multicolumn{4}{c}{\textbf{FashionMNIST (in) / MNIST (out)}} \\
             \midrule
             \multicolumn{4}{l}{\textbf{Use prior knowledge of OOD}} \\
    Backgr. contrast. LR (PixelCNN) {\parencite{ren_likelihood_2019}}               & $0.994$ & $0.993$ & $0.001$ \\
    Backgr. contrast. LR (VAE) {\parencite{choi_waic_2019}}                    & $0.924$ & - & - \\
    Binary classifier {\parencite{ren_likelihood_2019}}                              & $0.455$ & $0.505$ & $0.886$ \\ % 6
    $p(\hat{y} | \xb)$ with OOD as noise class {\parencite{ren_likelihood_2019}}     & $0.877$ & $0.871$ & $0.195$ \\ % 7
    $p(\hat{y} | \xb)$ with calibration on OOD {\parencite{ren_likelihood_2019}}     & $0.904$ & $0.895$ & $0.139$ \\ % 8
    Input complexity ($S$, Glow) \parencite{hendrycks_deep_2019}                    & $0.998$ & - & - \\
    Input complexity ($S$, PixelCNN++) \parencite{hendrycks_deep_2019}              & $0.967$ & - & - \\
             \multicolumn{4}{l}{\textbf{Use in-distribution data labels $y$}} \\
    $p(\hat{y} | \xb)$ {\parencite{ren_likelihood_2019, hendrycks_baseline_2017}}                        & $0.734$ & $0.702$ & $0.506$ \\
    Entropy of $p(y | \xb)$ {\parencite{ren_likelihood_2019}}                        & $0.746$ & $0.726$ & $0.448$ \\
    ODIN {\parencite{ren_likelihood_2019, liang_enhancing_2018}}                                       & $0.752$ & $0.763$ & $0.432$ \\
    VIB \parencite{alemi_uncertainty_2018, choi_waic_2019}                                          & $0.941$ & - & - \\
    Mahalanobis distance, CNN {\parencite{ren_likelihood_2019}}                     & $0.942$ & $0.928$ & $0.088$ \\
    Mahalanobis distance, DenseNet {\parencite{lee_simple_2018}}                & $0.986$ & - & - \\
    Ensemble, 20 classifiers {\parencite{ren_likelihood_2019, lakshminarayanan_simple_2017}}                  & $0.857$ & $0.849$ & $0.240$ \\
             \multicolumn{4}{l}{\textbf{No OOD-specific assumptions}} \\
             \multicolumn{4}{l}{\textit{- Ensembles}} \\
    WAIC, 5 models, VAE {\parencite{choi_waic_2019}}                          & $0.766$ & - & - \\
    WAIC, 5 models, PixelCNN {\parencite{ren_likelihood_2019}}                      & $0.221$ & $0.401$ & $0.911$ \\
            \multicolumn{4}{l}{\textit{- Not ensembles}} \\
    Likelihood regret \parencite{xiao_likelihood_2020}                               & $\mathbf{0.988}$ & - & - \\
    $\mathcal{L}^{>0}$ + HVAE (ours)                    & $0.268$ & $0.363$ & $0.882$ \\
    $\mathcal{L}^{>1}$ + HVAE (ours)                    & $0.593$ & $0.591$ & $0.658$ \\
    $\mathcal{L}^{>2}$ + HVAE (ours)                    & $0.712$ & $0.750$ & $0.548$ \\
    $LLR^{>1}$ + HVAE (ours)                            & $0.964$ & $0.961$ & $0.036$ \\
    $LLR^{>1}_{250}$ + HVAE (ours)                      & $0.984$ & $\mathbf{0.984}$ & $\mathbf{0.013}$ \\
             \bottomrule
        \end{tabular}%
        }
    \end{table}
\end{frame}

\begin{frame}
    \frametitle{Results on CIFAR10/SVHN}
    \begin{table}
        \centering
        \resizebox*{!}{0.87\textheight}{%
        \begin{tabular}{lrrr}
            \toprule
             Method & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
             \midrule
             \multicolumn{4}{c}{\textbf{CIFAR10 (in) / SVHN (out)}} \\
             \midrule
             \multicolumn{4}{l}{\textbf{Use prior knowledge of OOD}} \\
    Backgr. contrast. LR (PixelCNN) {\parencite{ren_likelihood_2019}}               & $0.930$ & $0.881$ & $0.066$ \\
    Backgr. contrast. LR (VAE) {\parencite{xiao_likelihood_2020}}                    & $0.265$ & - & - \\
    Outlier exposure {\parencite{hendrycks_deep_2019}}                              & $0.984$ & - & - \\
    Input complexity ($S$, Glow) \parencite{serra_input_2020}                   & $0.950$ & - & - \\
    Input complexity ($S$, PixelCNN++) \parencite{serra_input_2020}             & $0.929$ & - & - \\
    Input complexity ($S$, HVAE) (Ours) \parencite{serra_input_2020}\textsuperscript{\ref{footnote:serra_hierarchical_likelihood}} & $0.833$ & $0.855$ & $0.344$ \\
    \multicolumn{4}{l}{\textbf{Use in-distribution data labels $y$}} \\
    Mahalanobis distance {\parencite{lee_simple_2018}}                          & $0.991$ & - & -  \\
             \multicolumn{4}{l}{\textbf{No OOD-specific assumptions}} \\
             \multicolumn{4}{l}{\textit{- Ensembles}} \\
    WAIC, 5 models, Glow {\parencite{choi_waic_2019}}                          & $1.000$ & - & - \\
    WAIC, 5 models, PixelCNN {\parencite{ren_likelihood_2019}}                      & $0.628$ & $0.616$ & $0.657$ \\
             \multicolumn{4}{l}{\textit{- Not ensembles}} \\
    Likelihood regret \parencite{xiao_likelihood_2020}                               & $0.875$ & - & - \\
    $LLR^{>2}$ + HVAE (ours)                            & $0.811$ & $0.837$ & $0.394$ \\
    $LLR^{>2}$ + BIVA (ours)                            & $\mathbf{0.891}$ & $\mathbf{0.875}$ & $\mathbf{0.172}$ \\
             \bottomrule
        \end{tabular}%
        }
    \end{table}
\end{frame}

\frame{
    \frametitle{Results on diverse datasets}
    \begin{columns}
    \begin{column}{0.5\textwidth}

        \begin{table}[t]
            \centering
            \resizebox{0.9\textwidth}{!}{%
            \begin{tabular}{llrrr}
                \toprule
                 OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on CIFAR10}} \\
                 \midrule
        SVHN          &  $LLR^{>2}$  &  0.811  &  0.837  &  0.394  \\
        CIFAR10       &  $LLR^{>1}$  &  0.469  &  0.479  &  0.835  \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on SVHN}} \\
                 \midrule
        CIFAR10            &  $LLR^{>1}$       &  $0.939$  &  $0.950$  &  $0.052$  \\
        SVHN               &  $LLR^{>1}$       &  $0.489$  &  $0.484$  &  $0.799$  \\
            \bottomrule
            \end{tabular}
            }
        \end{table}
    \end{column}
    \begin{column}{0.5\textwidth}
        \begin{table}[t]
            \centering
            \resizebox{0.9\textwidth}{!}{%
            \begin{tabular}{llrrr}
                \toprule
                 OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on FashionMNIST}} \\
                 \midrule
        MNIST                    & $LLR^{>1}$             &  0.986  &  0.987  &  0.011 \\
        notMNIST                 &  $LLR^{>1}$            &  0.998  &  0.998  &  0.000 \\
        KMNIST                   &  $LLR^{>1}$            &  0.974  &  0.977  &  0.017 \\
        Omniglot28x28            &  $LLR^{>2}$            &  1.000  &  1.000  &  0.000 \\
        Omniglot28x28Inverted    &  $LLR^{>1}$            &  0.954  &  0.954  &  0.050 \\
        SmallNORB28x28           &  $LLR^{>2}$            &  0.999  &  0.999  &  0.002 \\
        SmallNORB28x28Inverted   &  $LLR^{>2}$            &  0.941  &  0.946  &  0.069 \\
        FashionMNIST             &  $LLR^{>1}$            &  0.488  &  0.496  &  0.811 \\
                 \midrule
                 \multicolumn{5}{c}{\textbf{Trained on MNIST}} \\
                 \midrule
        FashionMNIST                   &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
        notMNIST                       &  $LLR^{>1}$  &  $1.000$  &  $0.999$  &  $0.000$ \\
        KMNIST                         &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
        Omniglot28x28                  &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
        Omniglot28x28Inverted          &  $LLR^{>1}$  &  $0.944$  &  $0.953$  &  $0.057$ \\
        SmallNORB28x28                 &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
        SmallNORB28x28Inverted         &  $LLR^{>1}$  &  $0.985$  &  $0.987$  &  $0.000$ \\
        MNIST                          &  $LLR^{>2}$  &  $0.515$  &  $0.507$  &  $0.792$ \\
                 \bottomrule
            \end{tabular}
            }
            % \caption{Additional results for the HVAE model trained on FashionMNIST. All results computed with 1000 importance samples.}
            % \label{tab:additional-results-fashionmnist}
        \end{table}
    \end{column}
